---
title: "Machine Learning - Barbell Lifting Data"
author: "James Henzi"
date: "Wednesday, March 18, 2015"
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---

### Summary
The goal of this project is to use a machine learning algorithm to generate a prediction model to predict the "classe"" variable from the Weight Lifting Exercise Dataset found at http://groupware.les.inf.puc-rio.br/har.  This report details the data exploration/preparation, model generation, model validation, and finally the testing results.


### Data upload, exploration, cleaning, and preparation

The csv files for the training and testing data is uploaded from the web. 
```{r,cache=TRUE}
training<-read.csv("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
testing<-read.csv("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```

A summary view of the training data (see appendix) shows that there are many parameters that provide little information useful for our prediction (very high percentage of NAs or blanks, date/time tags for measurments). Thus these parameters are removed.

```{r, cache=TRUE,results='hide'}
dataTrain<-training[,c(1:2,4,8:11,37:49,60:68,84:86,102,113:124,140,151:160)]
dataTest<-testing[,c(1:2,4,8:11,37:49,60:68,84:86,102,113:124,140,151:160)]
```

Since the testing data does not include the result we are trying to predict (classe), we will need to slice the training data to create a validation data set that will be used to perform an interim test of the prediction model.

```{r,warning=FALSE,message=FALSE,cache=TRUE}
library(caret)
set.seed(1234)
inValidation<-createDataPartition(y=dataTrain$classe,p=0.75,list=F)
train2<-dataTrain[inValidation,]
validation<-dataTrain[-inValidation,]
```

### Model Creation

Given that we are looking at a categorical classification based on a reasonably large number of parameters a tree type model would most likely be a good option.  Thus , it was decided to try a random forest model.

```{r,cache=TRUE}
set.seed(4321)
modRF<-train(classe~.-X,data=train2,method="rf", ntree=50) 
## X excluded as it is a column reference number
## ntree set to 50 to reduce processing time
```

### Cross Validation

After model generation we tested it out on the validation data set generated earlier.
```{r,warning=FALSE,message=FALSE}
library(caret)
confusionMatrix(validation$classe,predict(modRF,validation))
```

From the confusion matrix we find that the model accurately classifies 99.37%.  This yields an expected out of sample error rate of 0.63%.  Since we achieved such a high level of accuracy, we will go with this model.


### Testing

The model is then applied to the test data set to predict the classifiaction for these 20 samples.

```{r}
predict(modRF,dataTest)
```

These predictions were then entered into the "Submission" portion of the project.  All 20 were predicted accurately!!

### Appendix
#### Summary of training data

```{r}
summary(training)
```

